---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit README.Rmd, not README.md. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# forenseqhaplo

<!-- badges: start -->

<!-- Badges for R CMD check, package version, DOI and license will be added after repository publication. -->

<!-- badges: end -->

`forenseqhaplo` is an R package for processing ForenSeq Flanking Region Reports, assigning sequence-based iSNP haplotypes, and generating output files for downstream forensic genetic analyses with Familias.

The package reads one or multiple ForenSeq UAS Excel reports, applies configurable sequence-selection, read-depth, and allele-balance rules, matches the selected sequences against a bundled or user-provided haplotype database, and produces:

* Familias-compatible haplotype TXT files.
* Per-sample Excel review reports.
* Processing logs and parameter records.
* Optional TargetSNP allele TXT files.
* Optional DVI relationship columns.
* Consensus profiles from replicate analyses.

> [!IMPORTANT]
> `forenseqhaplo` is intended to support data processing and manual review. Users are responsible for validating the analytical thresholds, haplotype database, and generated results according to their own laboratory procedures and quality requirements.

## Installation

### From GitHub

After repository publication, the stable version can be installed with:

```r
install.packages("remotes")

remotes::install_github(
  "FISABIO-BioinformaticsService/forenseqhaplo",
  ref = "v1.0.0",
  dependencies = TRUE,
  upgrade = "never"
)
```

The latest development version can be installed from the main branch:

```r
install.packages("remotes")

remotes::install_github(
  "FISABIO-BioinformaticsService/forenseqhaplo",
  dependencies = TRUE,
  upgrade = "never"
)
```

Dependencies declared in the package `DESCRIPTION` file will be installed automatically when possible.

### From CRAN

After a future CRAN acceptance, the package will be installable with:

```r
install.packages("forenseqhaplo")
```

### From a local source package

If you have received a local source package such as `forenseqhaplo_1.0.0.tar.gz`, install it with:

```r
install.packages("remotes")

remotes::install_local(
  "path/to/forenseqhaplo_1.0.0.tar.gz",
  dependencies = TRUE,
  upgrade = "never"
)
```

This is the recommended method for installing the package on another computer because it also installs missing dependencies from CRAN when internet access is available.

A minimal local installer script can also be distributed together with the package source file:

```r
cran_repo <- "https://cloud.r-project.org"
options(repos = c(CRAN = cran_repo))

pkg_file <- list.files(
  path = ".",
  pattern = "^forenseqhaplo_.*\\.tar\\.gz$",
  full.names = TRUE
)

if (length(pkg_file) == 0) {
  stop("No forenseqhaplo_*.tar.gz file was found in the current folder.")
}

if (length(pkg_file) > 1) {
  pkg_file <- pkg_file[which.max(file.info(pkg_file)$mtime)]
  message("More than one package file was found. Using the most recent one: ", basename(pkg_file))
}

if (!requireNamespace("remotes", quietly = TRUE)) {
  install.packages("remotes", repos = cran_repo)
}

remotes::install_local(
  path = pkg_file,
  dependencies = TRUE,
  upgrade = "never"
)

message("Installation completed. You can now run: library(forenseqhaplo)")
```

## Quick start

The main workflow processes all valid ForenSeq Flanking Region Report files found in an input directory:

```{r folder-example, eval = FALSE}
library(forenseqhaplo)

result <- process_forenseq_folder(
  input_dir = "path/to/forenseq_reports",
  output_dir = "path/to/results"
)
```

By default, this creates:

```text
results/
├── haplotypes.txt
├── parameters.xlsx
├── processing_log.tsv
├── review_sample_1.xlsx
├── review_sample_2.xlsx
└── ...
```

The returned R object contains the combined Familias table, processing log, parameters, and output paths:

```{r result-example, eval = FALSE}
result$familias_txt
result$processing_log
result$processing_parameters
result$output_txt
result$output_parameters_xlsx
```

## Processing a single sample

A single ForenSeq report can be processed with `process_forenseq_sample()`:

```{r sample-example, eval = FALSE}
result <- process_forenseq_sample(
  report_xlsx = "path/to/sample_Flanking_Region_Report.xlsx",
  output_xlsx = "path/to/review_sample.xlsx",
  output_txt = "path/to/haplotypes.txt"
)
```

The returned object provides access to the final calls and the loci requiring review:

```{r sample-result-example, eval = FALSE}
result$final_calls
result$review
result$all_sequences
result$familias_row
result$processing_parameters
```

## Input requirements

### ForenSeq reports

By default, the package reads the `iSNP Coverage` sheet from a ForenSeq Flanking Region Report.

The sample identifier is read from cell `B5`. If this value is missing during folder processing, the input filename is used as a fallback.

The iSNP table must contain the following information:

| Required field | Accepted column names                                                             |
| -------------- | --------------------------------------------------------------------------------- |
| Locus          | `iSNP Locus`, `iSNP & Variant Reference SNP`, or `iSNP and Variant Reference SNP` |
| Detected base  | `Detected Bases` or `Detected Base`                                               |
| Read count     | `Read`, `Reads`, or `Read Count`                                                  |
| Sequence       | `Sequence`                                                                        |

The package automatically searches the selected sheet for the table header.

### Input folder

Folder processing considers `.xlsx` files and ignores:

* Temporary Excel files beginning with `~$`.
* Previously generated review files.
* Haplotype and target output files.
* Duplicate-database reports.
* Processing logs and parameter files.

## Default interpretation parameters

The default analytical parameters are:

| Parameter                          | Default | Purpose                                                                     |
| ---------------------------------- | ------: | --------------------------------------------------------------------------- |
| `heterozygote_threshold`           |    0.30 | General minimum ratio between the second and first most abundant sequences  |
| `min_homozygote_reads`             |      30 | Minimum reads required to duplicate a single sequence as a homozygous call  |
| `min_heterozygous_reads`           |      11 | Minimum reads required for the second selected sequence in a heterozygote   |
| `extra_signal_threshold`           |    0.25 | Minimum third-to-first sequence ratio used to flag additional signal        |
| `review_balance_lower_margin`      |    0.15 | Lower proportional margin used to flag calls close to the balance threshold |
| `review_balance_upper_margin`      |    0.15 | Upper proportional margin used to flag calls close to the balance threshold |
| `review_low_homozygote_multiplier` |    1.50 | Multiplier used to flag homozygous calls close to the minimum read threshold |

All minimum read and ratio comparisons are inclusive. A value exactly equal to its threshold is accepted.

Some loci use lower default allele-balance thresholds:

| Threshold | Loci                                  |
| --------: | ------------------------------------- |
|      0.10 | `rs729172`, `rs10776839`, `rs1335873` |
|      0.20 | `rs338882`, `rs1493232`, `rs6955448`  |

All other loci use the general threshold of `0.30`.

The current defaults can be inspected directly:

```{r defaults-example, eval = FALSE}
default_heterozygote_threshold()
default_thresholds_by_locus()
default_min_homozygote_reads()
default_min_heterozygous_reads()
default_extra_signal_threshold()
default_review_balance_lower_margin()
default_review_balance_upper_margin()
default_review_low_homozygote_multiplier()
```

## Interpretation overview

For each locus, observed sequences are ordered by decreasing read count.

The main decision rules are:

1. If only one sequence is observed and its read count is at least `min_homozygote_reads`, the sequence is duplicated as a homozygous call.

2. If at least two sequences are observed, the allele balance is calculated as:

   ```text
   read count of sequence 2 / read count of sequence 1
   ```

3. The two most abundant sequences are selected as a heterozygous candidate when:

   * the allele balance is equal to or greater than the applicable threshold; and
   * the second sequence has at least `min_heterozygous_reads` reads.

4. If the balance is below the applicable threshold but the primary sequence reaches the homozygous read threshold, the primary sequence is duplicated and the locus is flagged for review.

5. Calls that do not reach the required read thresholds are reported as not interpretable.

6. A third sequence is flagged for review when its ratio relative to the primary sequence reaches `extra_signal_threshold`.

7. Selected sequences are matched against the haplotype database using the normalized `TargetSNP + Sequence` key.

The complete scientific justification and all review rules are described in the package documentation and associated publication.

## Output files

### `haplotypes.txt`

Wide, tab-separated file containing one row per sample and one column per locus.

The first column is `sample_id`. Each populated locus contains two comma-separated haplotypes:

```text
sample_id    rs10495407    rs1294331    ...
Sample_001   H1,H1         H2,H3         ...
```

### Per-sample review Excel files

Each processed sample generates an Excel workbook named:

```text
review_<sample_id>.xlsx
```

It contains three sheets:

| Sheet           | Content                                                     |
| --------------- | ----------------------------------------------------------- |
| `Final_calls`   | Final interpretation and haplotypes selected for each locus |
| `Review`        | Loci and observations requiring manual review               |
| `All_sequences` | All sequences observed in the original report               |

Analytical parameters are not repeated in the Excel review workbooks. They are recorded separately in `parameters.xlsx`.

### `processing_log.tsv`

Folder processing generates a summary containing:

* Processing status.
* Error messages.
* Original and final sample identifiers.
* Number of loci processed.
* Number of homozygous, heterozygous, and no-call loci.
* Number of review records.
* Number of database-matching failures.

By default, an error in one sample does not stop the remaining samples. This behaviour can be changed with:

```{r stop-on-error-example, eval = FALSE}
process_forenseq_folder(
  input_dir = "path/to/forenseq_reports",
  output_dir = "path/to/results",
  continue_on_error = FALSE
)
```

### `parameters.xlsx`

The parameters used during processing are recorded by default to support traceability and reproducibility.

Parameter export can be disabled with:

```{r no-parameters-example, eval = FALSE}
process_forenseq_folder(
  input_dir = "path/to/forenseq_reports",
  output_dir = "path/to/results",
  write_parameters_xlsx = FALSE
)
```

A custom parameter filename can be supplied with `output_parameters_xlsx`.

## Customising thresholds

All main thresholds can be modified:

```{r custom-thresholds-example, eval = FALSE}
result <- process_forenseq_folder(
  input_dir = "path/to/forenseq_reports",
  output_dir = "path/to/results",
  heterozygote_threshold = 0.30,
  min_homozygote_reads = 30,
  min_heterozygous_reads = 11,
  extra_signal_threshold = 0.25,
  review_balance_lower_margin = 0.15,
  review_balance_upper_margin = 0.15,
  review_low_homozygote_multiplier = 1.5
)
```

Locus-specific thresholds can be supplied as a named numeric vector:

```{r locus-thresholds-example, eval = FALSE}
custom_thresholds <- c(
  rs729172 = 0.12,
  rs338882 = 0.22
)

result <- process_forenseq_folder(
  input_dir = "path/to/forenseq_reports",
  output_dir = "path/to/results",
  thresholds_by_locus = custom_thresholds
)
```

Loci absent from `thresholds_by_locus` use the general `heterozygote_threshold`.

## Haplotype database

The package contains a bundled haplotype database used by default.

It can be inspected in R:

```{r database-example, eval = FALSE}
haplotype_db <- load_haplotype_database()

head(haplotype_db)
dim(haplotype_db)
```

It can also be exported to Excel:

```{r database-export-example, eval = FALSE}
export_haplotype_database(
  output_path = "haplotype_database.xlsx",
  overwrite = TRUE
)
```

A custom database must contain these columns:

| Column              | Description                                        |
| ------------------- | -------------------------------------------------- |
| `TargetSNP`         | Target iSNP identifier                             |
| `TargetSNP_allele`  | Allele at the target SNP                           |
| `Haplotype`         | Haplotype identifier used in output files          |
| `Sequence`          | Flanking-region sequence                           |
| `Polymorphic_sites` | Polymorphic positions represented by the haplotype |
| `Position_GRCh38`   | Genomic position according to GRCh38               |

The effective matching key is the combination of `TargetSNP` and normalized `Sequence`.

A custom database can be validated before processing:

```{r database-validation-example, eval = FALSE}
validate_haplotype_database(
  haplotype_db = "path/to/custom_haplotype_database.xlsx"
)
```

It can then be used during analysis:

```{r custom-database-example, eval = FALSE}
result <- process_forenseq_folder(
  input_dir = "path/to/forenseq_reports",
  output_dir = "path/to/results",
  haplotype_db = "path/to/custom_haplotype_database.xlsx"
)
```

Duplicated `TargetSNP + Sequence` keys are treated as invalid because they produce ambiguous sequence matching.

## Optional TargetSNP allele output

In addition to the haplotype TXT, the package can generate a file containing the target SNP alleles:

```{r target-example, eval = FALSE}
result <- process_forenseq_folder(
  input_dir = "path/to/forenseq_reports",
  output_dir = "path/to/results",
  write_target_txt = TRUE
)
```

This creates:

```text
target.txt
```

A different filename can be provided with `output_target_txt`.

## DVI mode

DVI mode adds relationship information to the generated TXT output.

The relationships Excel file must contain:

| Column         | Description                                        |
| -------------- | -------------------------------------------------- |
| `sample_id`    | Sample identifier matching the final TXT sample ID |
| `relationship` | Relationship category                              |
| `family_unit`  | Family or DVI unit identifier                      |

Example:

```{r dvi-example, eval = FALSE}
result <- process_forenseq_folder(
  input_dir = "path/to/forenseq_reports",
  output_dir = "path/to/results",
  dvi = TRUE,
  dvi_relationships = "path/to/dvi_relationships.xlsx"
)
```

Each `sample_id` must occur exactly once in the relationships file.

## Replicate consensus

A consensus TXT can be generated from multiple review Excel files produced by the package:

```{r consensus-example, eval = FALSE}
consensus <- create_consensus_txt(
  reports_dir = "path/to/replicate_review_files",
  output_txt = "consensus_haplotypes.txt"
)
```

Potential conflicts and missing loci can be inspected with:

```{r consensus-results-example, eval = FALSE}
consensus$conflicts
consensus$missing_loci
consensus$consensus_txt
```

## Reproducibility

For reproducible analyses, record at least:

* The `forenseqhaplo` version.
* The haplotype database version.
* All non-default thresholds.
* The ForenSeq UAS version used to generate the reports.
* The processing date.
* The reference genome assembly used by the haplotype database.
* Any manual review or correction applied after automatic processing.

The installed package version can be obtained with:

```r
packageVersion("forenseqhaplo")
```

## Documentation

The complete documentation will be available at:

```text
https://fisabio-bioinformaticsservice.github.io/forenseqhaplo/
```

Individual function help pages are also available from R:

```{r help-example, eval = FALSE}
?process_forenseq_folder
?process_forenseq_sample
?create_consensus_txt
?validate_haplotype_database
```

## Citation

If you use `forenseqhaplo` in a scientific publication, please cite both the software and the associated paper.

### Software

```text
Soriano Chirona V, et al. forenseqhaplo: Prepare ForenSeq haplotypes for
Familias. Version 1.0.0. Zenodo. DOI to be added after the first archived release.
```

### Associated publication

```text
The reference and DOI of the associated publication will be added after publication.
```

The definitive citation will be available through:

```r
citation("forenseqhaplo")
```

## Data confidentiality

Do not publish or share ForenSeq reports containing identifiable forensic or genetic information unless their use and distribution have been specifically authorised.

Input reports, review workbooks, relationship files, and generated TXT files should be managed according to the applicable institutional, legal, ethical, and information-security requirements.

## Issues and contributions

Bug reports and feature requests can be submitted through:

```text
https://github.com/FISABIO-BioinformaticsService/forenseqhaplo/issues
```

When reporting a problem, do not attach sensitive or identifiable forensic data. Use only fully synthetic examples that do not originate from a real individual, biological sample, sequencing run, or forensic case.

## License

`forenseqhaplo` is distributed under the GNU General Public License version 3 (`GPL-3`).

See the `LICENSE` file for details.
